Visual Odometry for Autonomous Underwater Vehicles

### 2013-03-26 paper.tex
Added the acknowlegment section
some corrections

\author{\IEEEauthorblockN{
Stephan Wirth, 
Pep Lluis Negre Carrasco, 
Gabriel Oliver-Codina}
\IEEEauthorblockA{
Departament de Matemàtiques i Informàtica, 
Universitat de les Illes Balears\\
Cra. Valldemossa, km. 7,5 07122 Palma de Mallorca (Spain)\\
Email: \{stephan.wirth, pl.negre, goliver\}@uib.es}}

@@

The lastest advances in Autonomous Underwater Vehicles (AUVs) demand accurate estimations of the vehicle's pose and velocity. The tasks to be performed by such robots are increasingly complex and, in most cases it is needed to precisely positioning the underwater vehicle to a certain location in order to execute the desired task. Navigation systems based on wirelessly transmited data, such as Global Positioning System (GPS) or mobile network positioning, can not be used in uderwater enviroments. Therefore the AUV must estimate its position using self-contained sensors. One of the most popular positioning sensors for underwater vehicles is the Doppler Velocity Log (DVL) which provides precise velocity and altitude updates for a wide variety of underwater tasks. However, DVL is an expensive device that cannot be integraded in many underwater vehicles.

Visual odometry is the process of estimating a vehicle 3D pose using only visual images. This technique is becoming popular in AUVs for navigation, station keeping and the provision of a manipulator feedback information. Visual odometry output is often fused with Inertial Measurement Units (IMU) to provide a cheaper alternative to DVLs \cite{Hildebrandt2010} \cite{Dunbabin2004}. However, the limitations of underwater vision are widely known, and its performance depends on many factors such as visibility, lighting and distorsion resulting from varying refractive indices.

In this paper we present the integration of two existing stereo visual odometry algorithms into an AUV, the experiments carried out in laboratory and harbour enviroments, as well as the evaluation method to compare the performance of both algorithms.

First, the paper explains related works on visual odometry. Section \ref{visual-odometry-systems} describes the general concepts of visual odometry systems and presents a brief overview of the two analyzed algorithms (fovis and libviso2), followed by the explanation of the evaluation method in Section \ref{experiments} and experiments in different eviroments. Finally, the conclusion of the study and future prespectives are presented.

@@

The basic visual odometry algorithm pipeline \cite{Moravec1980} consists, first, in identifying keypoints (landmarks) in each camera frame and extract the feature descriptors of these points. Then compute the depth for every landmark (typically using stereo or rgb-d cameras), match features across time frames and estimate the rigid-body transformation that best aligns the features between frames. The result of this process is an estimation of  camera motion between frames and, therefore it is necessary to integrate this data over the time to obtain the vehicle position and orientation.

Visual odometers for AUVs tend to use this pipeline \cite{Botelho2009} and adapt every stage options (such us keypoint detector type and feature matching method) to their needs. Common feature detectors used for real-time visual odometry include Harris corners \cite{Harris1988,Nister2006}, FAST \cite{Rosten2006,Huang2011} or SIFT \cite{Lowe2004,Botelho2009} features, which have a low computation cost and are robust against small viewpoint changes. Wide range of approaches are involved in the process of matching feature across frames, \cite{Geiger2011} propose a circle match between stereo image pairs of two consecutive frames and uses non-minima/maxima-suppression techniques \cite{Geiger2010} to reduce the number of correspondences. RANSAC-based methods \cite{Nister2004,Johnson2008} and graph-based consistency algorithms \cite{Howard2008} have also been proven to robustly match features across frames.

Previous to the motion estimation stage, keypoint bucketing \cite{Zhang1995} has been shown to help reduce the inlier reprojection errors. Finally, motion estimation process can be solved through different methods. Using the closed-form solution to least-squares problem of absolute orientation \cite{Horn1987}, it is possible to directly minimize the Euclidean distance between matched features in order to compute the rigid-body transformation of the camera frame \cite{Huang2011}. Several ground-based visual odometers do not use 3D distances but implement the method of minimizing the pixel reprojection error \cite{Howard2008,Geiger2011}.

Other AUVs localization algorithms are based on mosaics or structured enviroments and therefore assume prior knowledge of the area to explore \cite{Garcia2001,Gracias2003,Carreras2003}.

Many of the techniques and algorithms described in this section are primary focused on ground and aerial vehicles, however most of them have not been tested in underwater environments. In this field, the complexity of the visual odometry task increases due to the dynamic light conditions and decreasing visibility with depth and turbity. Our focus is primarily on integrating and validating two different stereo-based visual odometry approaches in such enviroments.

@@

We created wrappers for both libraries\footnote{See \url{http://www.ros.org/viso2_ros} and \url{http://www.ros.org/fovis_ros} for source code and documentation.} to integrate them into the Robot Operating System ROS \cite{Quigley2009}. This simplifies the integration in different robotic vehicles as well as comparison with other integrated sensors.

Both algorithms have very similar processing pipelines containing feature detection, filtering, matching, and motion estimation. Table \ref{odometry-comparison} summarizes the steps of the algorithms, outlining similarities and differences. To limit the extend of this paper, the interested reader is refered to the original papers for a deeper understanding and more detailed description of the algorithms. In the following we give a brief overview of both algorithms.

@@

Subsequently, 3D coordinates of the features in the previous image pair are calculated through triangulation. Gauss-Newton minimization of the reprojection error of these 3D points onto the current left and right images leads to the rigid transformation from the previous image's camera pose to the current camera pose. The motion estimation procedure is wrapped in a RANSAC \cite{Fischler1981} scheme.

@@

fovis uses the FAST feature detector \cite{Rosten2005FPA,Rosten2006} with an adaptive threshold on three gaussian pyramid levels. The number of features is reduced taking the best features of each cell of a regular grid. The feature descriptor is an intensity normalized 9x9 pixel window centered on the feature location. The bottom right pixel is left out to get better memory alignment for rapid similarity measure computation using SIMD instructions.

@@

\begin{table*}[ht]
  \caption{Comparison of the two visual odometry algorithms.}
  \label{odometry-comparison}
  \begin{tabularx}{\linewidth}{l|X|X}
    & \textbf{libviso2} & \textbf{fovis}  \\
    \hline
    Feature Scales                  & 1 (reduced) level & 3 gaussian pyramid levels \\[3pt]
    Feature Detection               & 5x5 blob and corner masks & FAST with adaptive threshold  \\[3pt]
    Initial Feature Reduction       & non-minimum-/non-maximum-suppression & grid filter \\[3pt]
    Descriptor                      & 16 Sobel filter responses & 9x9 pixel window, intensity normalized \\[3pt]
    Subpixel Position Refinement    & parabolic fitting & ESM \\[3pt]
    \hline
    Matching Search Space Limitation & two-step candidate reduction & initial rotation estimate (Mei et al.) + search window \\[3pt]
    Matching Images                 & circlular match & left to right \& left to previous left \\[3pt]
    Match Reduction                 & grid filter & none \\[3pt]
    \hline
    Initial Outlier Rejection       & 2D neighborhood support & none \\[3pt]
    Outlier Classification          & iteratively during motion estimation (RANSAC) & maximally consistent clique \\[3pt]
    Minimized Reprojection Error    & previous 3D to current left and right image & previous 3D to current left and current 3D to previous left image \\[3pt]
    \hline
    Motion Filtering                & Kalman & None (external) \\[3pt]
    Keyframe Selection              & none (external) & automatic, based on number of matches \\[3pt]
  \end{tabularx}
\end{table*}

@@

\begin{itemize}
  \item Two Bumblebee2 stereo cameras from Point Grey, one with a focal distance of 3.8mm (66\degree HFOV), and the other with 2.5mm (97\degree HFOV).
  \item A mini-ITX motherboard from Jetway-NF98-QM57 Express Chipset with an Intel i5 processor at 2.33GHz with 4 cores.
  \item A PCI express card with two firewire ports from Chronos to connect the cameras to the motherboard.
\end{itemize}

@@

For the harbour experiment, the determination of ground truth is more difficult, as no external sensors have been used and positions of natural landmarks are not known. One aim of the project was the construction of a consistent seabed mosaic (see \cite{Ferrer2007}). The computation of this mosaic includes global optimization of all camera poses. The resulting trajectory does not suffer from drift and is therefore chosen as our ground truth reference.

@@

A comparison of linear velocity estimates to ground truth can be found in Figure \ref{error-plots-viso2} for libviso2 and \ref{error-plots-fovis} for fovis. Upper graphics refer to experiments in laboratory while the ones of the bottom refer to harbour tests. Linear velocity is shown in left graphics for both odometry estimation (blue) and ground truth (green), together with its error drawn in red. Finally, on right plot, a comparison of angular velocity for odometry estimation (blue) and grund truth (green) is also presented. The error in angular velocity has a zero mean thus not shown in the figures. No differences can be highlighted comparing angular velocity error for libviso2 and fovis in both the laboratory and harbour conditions. However, when linear velocities are examined, fovis presents better performance when executed in laboratory enviroment.

These results are numerically summarized in Table \ref{results-errors} for laboratory and harbour experiments. Translation column indicates the mean average translation error that suffers the odometer in percent, which is the number of meters diverted for every 100 meters of trajectory. Furthermore, the rotation column indicates how many degrees the odometer is deviated per meter, relative to ground truth.

@@

>>>>

\author{\IEEEauthorblockN{
Stephan Wirth, 
Pep Lluis Negre Carrasco, 
Gabriel Oliver Codina}
\IEEEauthorblockA{
Departament de Matemàtiques i Informàtica, 
Universitat de les Illes Balears\\
Cra. Valldemossa, km. 7,5 07122 Palma de Mallorca (Spain)\\
Email: \{stephan.wirth, pl.negre, goliver\}@uib.es}}

@@

The latest advances in Autonomous Underwater Vehicles (AUVs) demand accurate estimations of the vehicle's pose and velocity. The tasks to be performed by such robots are increasingly complex and, in most cases it is needed to precisely position the underwater vehicle to a certain location in order to execute the desired task. Navigation systems based on wirelessly transmitted data, such as the Global Positioning System (GPS) or mobile network positioning, can not be used in underwater environments. One can overcome this issue by introducing artificial landmarks. However the cost and effort for this is high and for small missions it is preferred to rely on positioning using self-contained sensors. One of the most popular positioning sensors for underwater vehicles is the Doppler Velocity Log (DVL) which provides precise velocity and altitude updates. However, it is an expensive device that cannot be integrated in many underwater vehicles.

Visual odometry is the process of estimating a vehicle's 3D pose using only visual images. This technique is becoming popular in AUVs for navigation, station keeping and the provision of feedback information for manipulation. Visual odometry output is often fused with Inertial Measurement Units (IMU) to provide a cheaper alternative to DVLs \cite{Hildebrandt2010,Dunbabin2004}. However, the limitations of underwater vision are widely known, and its performance depends on many factors such as visibility, lighting and distortion resulting from varying refractive indices.

In this paper we present the integration of two existing stereo visual odometry algorithms into an AUV, the experiments carried out in laboratory and harbour environments, as well as the evaluation method to compare the performance of both algorithms.

First, the paper explains related works on visual odometry. Section \ref{visual-odometry-systems} describes the general concepts of visual odometry systems and presents a brief overview of the two analyzed algorithms, followed by the explanation of the evaluation method in Section \ref{experiments} and experiments in different evironments. Finally, the conclusion of the study and future perspectives are presented.

@@

The basic visual odometry algorithm pipeline \cite{Moravec1980} consists of the following steps: First, keypoints (landmarks) are identified in each camera frame and feature descriptors for these points are extracted. Then, the depth for every landmark is estimated using stereo, structure from motion or a separate depth camera. Subsequently, features are matched across time frames and the rigid-body transformation that best aligns the features between frames is estimated. The result of this process is an estimation of camera motion between frames and therefore it is necessary to integrate this data over time to obtain the vehicle's absolute position and orientation.

Many implementations of visual odometers for AUVs use this pipeline and adapt every stage (such us keypoint detector type and feature matching method) to their needs. Common feature detectors used for real-time visual odometry include Harris corners \cite{Harris1988,Nister2006}, FAST \cite{Rosten2006,Huang2011} or SIFT \cite{Lowe2004,Botelho2009} features. Wide range of approaches are involved in the process of matching feature across frames, \cite{Geiger2011} propose a circle match between stereo image pairs of two consecutive frames and uses non-minima/maxima-suppression techniques \cite{Geiger2010} to reduce the number of correspondences. RANSAC-based methods \cite{Nister2004,Johnson2008} and graph-based consistency algorithms \cite{Howard2008} have also been proven to robustly match features across frames.

Previous to the motion estimation stage, keypoint bucketing \cite{Zhang1995} has been shown to help reduce the inlier reprojection errors. Finally, the motion estimation process can be solved through different methods. Using a closed-form solution to the least-squares problem of absolute orientation \cite{Horn1987}, it is possible to directly minimize the Euclidean distance between matched features in order to compute the rigid-body transformation of the camera frame \cite{Huang2011}. Several ground-based visual odometers do not use 3D distances but implement the method of minimizing the pixel reprojection error \cite{Howard2008,Geiger2011,Huan2011}.

Other localization algorithms are for AUVs are based on mosaics or structured environments and therefore assume prior knowledge of the area to explore \cite{Garcia2001,Gracias2003,Carreras2003}.

Many of the techniques and algorithms described in this section are primary focused on ground and aerial vehicles and most of them have not been tested in underwater environments. In this field, the complexity of the visual odometry task increases due to the dynamic light conditions and decreasing visibility with depth and turbidity. Our focus is primarily on integrating and validating two different stereo-based visual odometry approaches in such environments.

@@

We created wrappers for both libraries\footnote{See \url{http://www.ros.org/wiki/viso2_ros} and \url{http://www.ros.org/wiki/fovis_ros} for source code and documentation.} to integrate them into the Robot Operating System ROS \cite{Quigley2009}. This simplifies the integration in different robotic vehicles as well as comparison with other integrated sensors.

Both algorithms have very similar processing pipelines containing feature detection, filtering, matching, and motion estimation. Table \ref{odometry-comparison} summarizes the steps of the algorithms, outlining similarities and differences. To limit the extent of this paper, the interested reader is referred to the original papers for a deeper understanding and a more detailed description of the algorithms. In the following we give a brief overview of both algorithms.

@@

Subsequently, 3D coordinates of the features in the previous image pair are calculated through triangulation. Gauss-Newton minimization of the reprojection error of these 3D points onto the current left and right images leads to the rigid transformation from the previous image's camera pose to the current camera pose. The motion estimation procedure is wrapped in a RANSAC \cite{Fischler1981} scheme to get rid of outliers.

@@

fovis uses the FAST feature detector \cite{Rosten2005FPA,Rosten2006} with an adaptive threshold on three Gaussian pyramid levels. The number of features is reduced taking the best features of each cell of a regular grid. The feature descriptor is an intensity normalized 9x9 pixel window centered on the feature location. The bottom right pixel is left out to get better memory alignment for rapid similarity measure computation using SIMD instructions.

@@

\begin{table*}[ht]
  \caption{Comparison of the two visual odometry algorithms.}
  \label{odometry-comparison}
  \begin{tabularx}{\linewidth}{l|X|X}
    & \textbf{libviso2} & \textbf{fovis}  \\
    \hline
    Feature Scales                  & 1 (reduced) level & 3 gaussian pyramid levels \\[3pt]
    Feature Detection               & 5x5 blob and corner masks & FAST with adaptive threshold  \\[3pt]
    Initial Feature Reduction       & non-minimum-/non-maximum-suppression & grid filter \\[3pt]
    Descriptor                      & 16 Sobel filter responses & 9x9 pixel window, intensity normalized \\[3pt]
    Subpixel Position Refinement    & parabolic fitting & ESM \\[3pt]
    \hline
    Matching Search Space Limitation & two-step candidate reduction & initial rotation estimate (Mei et al.) + search window \\[3pt]
    Matching Images                 & circlular match & left to right \& left to previous left \\[3pt]
    Match Reduction                 & grid filter & none \\[3pt]
    \hline
    Initial Outlier Rejection       & 2D neighborhood support & none \\[3pt]
    Outlier Classification          & iteratively during motion estimation (RANSAC) & maximally consistent clique \\[3pt]
    Minimized Reprojection Error    & previous 3D to current left and right image & previous 3D to current left and current 3D to previous left image \\[3pt]
    \hline
    Motion Filtering                & Kalman & none (external) \\[3pt]
    Keyframe Selection              & none (external) & automatic, based on number of matches \\[3pt]
  \end{tabularx}
\end{table*}

@@

\begin{itemize}
  \item Two Point Grey Bumblebee2 stereo cameras, one with a focal distance of 3.8mm (66\degree HFOV), and the other with 2.5mm (97\degree HFOV). % HFOV in water?
  \item A mini-ITX motherboard from Jetway-NF98-QM57 Express Chipset with an Intel i5 processor at 2.33GHz with 4 cores.
  \item A PCI express card with two firewire ports from Chronos to connect the cameras to the motherboard.
\end{itemize}

@@

For the harbour experiment, the determination of ground truth is more difficult, as no external sensors have been used and positions of natural landmarks are not known. One aim of the project was the offline construction of a consistent seabed mosaic (see \cite{Ferrer2007}). The computation of this mosaic includes global optimization of all camera poses. The resulting trajectory does not suffer from drift and is therefore chosen as our ground truth reference.

@@

A comparison of linear velocity estimates to ground truth can be found in Figure \ref{error-plots-viso2} for libviso2 and \ref{error-plots-fovis} for fovis. Upper graphics refer to experiments in laboratory while the ones of the bottom refer to harbour tests. Linear velocity is shown in left graphics for both odometry estimation (blue) and ground truth (green), together with its error drawn in red. Finally, on right plot, a comparison of angular velocity for odometry estimation (blue) and grund truth (green) is also presented. No differences can be highlighted comparing angular velocity error for libviso2 and fovis in both the laboratory and harbour conditions. However, when linear velocities are examined, fovis presents better performance when executed in laboratory enviroment.

These results are summarized in Table \ref{results-errors} for laboratory and harbour experiments. The translation column indicates the mean average translation error that suffers the odometer in percent, which is the number of meters diverted for every 100 meters of trajectory. Furthermore, the rotation column indicates how many degrees the odometer is deviated per meter, relative to ground truth.

@@

\section{Acknowledgment
  \label{acknowledgment}
}

This work is partially supported by the Spanish Ministry of Economy and Competitiveness under contract DPI2011-27977-C03- 02 (TRITON Project), FEDER Fundings and by the European Commissions FP7 under grant agreement 248497 (TRIDENT Project).
